---
title: "Commute EDA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.asp = 0.618,
  out.width = "70%",
  fig.align = "center"
)
```

```{r include = FALSE}
library(tidyverse)
library(hms)
library(commuter)

data("chicago_to_harvard_monday_through_friday")
data("harvard_to_chicago_monday_through_friday")
```

## Trains between Arlington Heights and Ogilvie stations

```{r echo=FALSE}
morning_trains <- harvard_to_chicago_monday_through_friday %>%
  arrange(`Arlington Heights`) %>%
  mutate(train_number = factor(train_number, ordered = TRUE)) %>%
  gather(station, time, Ogilvie, `Arlington Heights`)

morning_trains %>%
  ggplot(aes(as.POSIXct(time), rev(train_number))) +
  geom_line() +
  geom_point() +
  scale_x_datetime() +
  labs(x = NULL, y = "Train #") +
  ggtitle("Morning inbound trains from Arlington Heights to Ogilvie") +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank()
  )
```

```{r echo=FALSE}
evening_trains <- chicago_to_harvard_monday_through_friday %>%
  arrange(Ogilvie) %>%
  mutate(train_number = factor(train_number, ordered = TRUE)) %>%
  gather(station, time, Ogilvie, `Arlington Heights`)

evening_trains %>%
  ggplot(aes(as.POSIXct(time), rev(train_number))) +
  geom_line() +
  geom_point() +
  scale_x_datetime() +
  labs(x = NULL, y = "Train #") +
  ggtitle("Evening outbound trains from Ogilvie to Arlington Heights") +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank()
  )
```

## Routes in and out

```{r echo=FALSE}
morning_options <- harvard_to_chicago_monday_through_friday %>%
  rename(morning_train_number = train_number,
         morning_depart = `Arlington Heights`,
         morning_arrive = Ogilvie)

evening_options <- chicago_to_harvard_monday_through_friday %>%
  rename(evening_train_number = train_number,
         evening_depart = Ogilvie,
         evening_arrive = `Arlington Heights`)

combined_options <- crossing(
  morning_options,
  evening_options
) %>%
  mutate(time_on_morning_train = as.hms(morning_arrive - morning_depart),
         time_on_evening_train = as.hms(evening_arrive - evening_depart),
         total_time_on_train =  as.hms(time_on_morning_train + time_on_evening_train))
```

```{r echo=FALSE}
combined_options %>%
  select(morning_train_number, morning_depart, morning_arrive, time_on_morning_train) %>%
  mutate(time_on_morning_train = as.numeric(time_on_morning_train) / 60) %>%
  distinct() %>%
  arrange(time_on_morning_train) %>%
  knitr::kable(col.names = c("Train number", "Depart", "Arrive", "Time on train (min)"))
```

- 12 trains in the morning
- 35--53 minutes on train

```{r echo=FALSE}
combined_options %>%
  select(evening_train_number, evening_depart, evening_arrive, time_on_evening_train) %>%
  distinct() %>%
  mutate(time_on_evening_train = as.numeric(time_on_evening_train) / 60) %>%
  arrange(time_on_evening_train) %>%
  knitr::kable(col.names = c("Train number", "Depart", "Arrive", "Time on train (min)"))
```

- 7 trains in the evening
- 36--54 minutes on train

```{r echo=FALSE}
combined_options %>%
  ggplot(aes(as.POSIXct(morning_depart), as.numeric(total_time_on_train) / 60 / 60)) +
  geom_hline(yintercept = 1, color = "gray", linetype = 2) +
  geom_hline(yintercept = 1.5, color = "gray", linetype = 2) +
  geom_hline(yintercept = 2, color = "gray", linetype = 2) +
  geom_jitter() +
  labs(title = "For 84 options, total time on train varies from 1.2 to 1.8 hours",
       subtitle = "The later you leave in the morning the better the options",
       x = "Time out on the morning train",
       y = "Total time on both trains (hours)") +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank()
  )
```

## Time in office

- Assuming 8-minute walk from station to office.

```{r echo=FALSE}
combined_options_in_office <- combined_options %>%
  mutate(office_arrive = hms::as.hms(morning_arrive + hms::hms(0, 8, 0)),
         office_depart = hms::as.hms(evening_depart - hms::hms(0, 8 + 20, 0)),
         total_time_in_office = hms::as.hms(office_depart - office_arrive))

combined_options_in_office %>%
  ggplot(aes(as.POSIXct(morning_depart), as.numeric(total_time_in_office) / 60 / 60)) +
  geom_hline(yintercept = 6, color = "gray", linetype = 2) +
  geom_hline(yintercept = 7, color = "gray", linetype = 2) +
  geom_hline(yintercept = 8, color = "gray", linetype = 2) +
  geom_jitter() +
  labs(title = "For 84 options, most are not viable",
       subtitle = "There's little benefit to leaving super early",
       x = "Time out on the morning train",
       y = "Total time in office (hours)") +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank()
  )
```

## Door-to-door

Assuming to/from station

1. 15 minute walk
2. 10 minute drive
3. 5 minute bike

```{r echo=FALSE}
combined_options_in_all <- combined_options_in_office %>%
  crossing(
    tribble(
      ~house_to_station_mode, ~house_to_station_time,
      "walk", hms(0, 15, 0),
      "drive", hms(0, 10, 0),
      "bike", hms(0, 5, 0)
    )
  ) %>%
  mutate(
    house_depart = as.hms(morning_depart - house_to_station_time),
    house_arrive = as.hms(evening_arrive + house_to_station_time)
  ) %>%
  mutate(
    route_id = as.character(row_number())
  )
```

Assessing options:

- Must be at house by 4:45 at the latest
- Must work in office at least 7 hours in office
- Must work in office no more than 8.25 hours

```{r echo=FALSE}
house_pickup_options <- combined_options_in_all %>%
  filter(house_arrive <= hms(0, 45, 16),
         total_time_in_office >= hms(0, 0, 7),
         total_time_in_office <= hms(0, 15, 8)) %>%
  mutate(total_time_in_office_level = ifelse(total_time_in_office < hms(0, 50, 6), "(-, 6:50)",
                                             ifelse(total_time_in_office < hms(0, 50, 7), "[6:50, 7:50)", "[7:50, -)"))) %>%
  arrange(house_depart, house_arrive, total_time_in_office) %>%
  mutate(route_id = factor(row_number(), ordered = TRUE))

house_pickup_station_to_house_time <- house_pickup_options %>%
  select(route_id, house_arrive) %>%
  mutate(house_arrive_label = stringr::str_trunc(as.character(as.hms(house_arrive - hms(0, 0, 12))), 5, side = "right", ellipsis = ""))

house_pickup_station_to_house <- house_pickup_options %>%
  select(route_id, house_to_station_mode, evening_arrive, house_arrive) %>%
  gather(location, time, evening_arrive, house_arrive)

house_pickup_office_to_train <- house_pickup_options %>%
  select(route_id, office_depart, evening_depart) %>%
  gather(location, time, office_depart, evening_depart)

house_pickup_evening_train <- house_pickup_options %>%
  select(route_id, evening_depart, evening_arrive) %>%
  gather(location, time, evening_depart, evening_arrive)

house_pickup_office_time <- house_pickup_options %>%
  select(route_id, total_time_in_office_level, office_arrive, office_depart) %>%
  gather(location, time, office_arrive, office_depart)

house_pickup_office_time_time <- house_pickup_options %>%
  select(route_id, total_time_in_office, total_time_in_office_level, office_arrive) %>%
  mutate(total_time_in_office_label = stringr::str_trunc(as.character(total_time_in_office), 5, side = "right", ellipsis = ""))

house_pickup_train_to_office <- house_pickup_options %>%
  select(route_id, morning_arrive, office_arrive) %>%
  gather(location, time, morning_arrive, office_arrive)

house_pickup_morning_train <- house_pickup_options %>%
  select(route_id, morning_depart, morning_arrive) %>%
  gather(location, time, morning_depart, morning_arrive)

house_pickup_house_to_train <- house_pickup_options %>%
  select(route_id, house_to_station_mode, house_depart, morning_depart) %>%
  gather(location, time, house_depart, morning_depart)

house_pickup_house_to_train_time <- house_pickup_options %>%
  select(route_id, house_depart) %>%
  mutate(house_depart_label = stringr::str_trunc(as.character(house_depart), 5, side = "right", ellipsis = ""))

ggplot() +
  geom_line(aes(as.hms(time), route_id, color = house_to_station_mode), house_pickup_house_to_train) +
  geom_line(aes(as.hms(time), route_id), house_pickup_morning_train, linetype = 2) +
  geom_line(aes(as.hms(time), route_id), house_pickup_train_to_office, linetype = 3) +
  geom_line(aes(as.hms(time), route_id), house_pickup_office_time) +
  geom_line(aes(as.hms(time), route_id), house_pickup_office_to_train, linetype = 3) +
  geom_line(aes(as.hms(time), route_id), house_pickup_evening_train, linetype = 2) +
  geom_line(aes(as.hms(time), route_id, color = house_to_station_mode), house_pickup_station_to_house) +
  geom_text(aes(house_arrive, route_id, label = house_arrive_label), house_pickup_station_to_house_time, hjust = 0) +
  geom_text(aes(house_depart, route_id, label = house_depart_label), house_pickup_house_to_train_time, hjust = 1) +
  geom_label(aes(office_arrive, route_id, label = total_time_in_office_label, color = total_time_in_office_level), house_pickup_office_time_time) +
  geom_vline(xintercept = hms(0, 15, 6), color = "gray", alpha = 1/2) +
  geom_vline(xintercept = hms(0, 30, 16), color = "gray", alpha = 1/2) +
  scale_x_time(limits = c(hms(0, 0, 5), hms(0, 0, 18))) +
  labs(x = NULL,
       y = NULL) +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        legend.position = "none",
        panel.grid = element_blank())
```

